---
title: "Day 3 Review and Lagging Variables"
author: "Rachel Mercaldo"
date: "2025-07-20"
output: html_document
---

Welcome to Day 3!

Let's start with a review of day 2.

On day 2, we covered an introduction to Weather Source data structure and primarily worked on merging Weather Source data files.
We got into a little bit of plotting.

Below is the live version of some of day 2's code (examples Rachel gave during a Zoom session). 

To begin, we need to load our libraries.

```{r}

library(dplyr)
library(stringr)
library(data.table)
library(tidyverse)

```


Next, we will load the data. 

Since the Day-2.Rmd file has all the code for arranging the data, today we will begin by loading the file "practice_data.csv". This file was generated using the code from day 2!

```{r}

data <- read.csv("practice_data.csv")

```


I went through a few examples of indexing on Day 1, and we did a little work on aggregating data and selecting certain variables from a dataset on Day 2. Below, I have a few more examples of a new function, *select()* that does the same things, but faster. This code also makes use of the %>% pipe operator.

You can read a command to R the same way you can read a sentence, and that helps make sense of the pipe operator. For example:

data %>% group_by(mon_year) %>% summarize(cases = sum(case_rate_100k))

can be read as:

"take dataset *data* and group the rows by the variable *mon_year*, then pipe the grouped rows to the *summarize()* function and calculate the sum of the case rates within each mon_year group."

You must also assign this to something for best results (otherwise, R will just print the grouped summaries). So really you want to use:

data2 <- data %>% group_by(mon_year) %>% summarize(cases = sum(case_rate_100k))

You saw me do some grouping and summarizing during the plotting steps of Day 2, though I wanted you to focus more on the GGPlot package at the time.


Let's use *select()*


```{r}

data2 <- data %>% 
  select(mon_year)

head(data2)

```

What happened? Data2 is now a new object in our Environment. It was created when we took the dataset *data* and asked R to pipe it to the function *select()*. That function selected only one variable, *mon_year*. 


We can also use a negative symbol within *select()* to *remove* variables we don't want. 

```{r}

data3 <- data %>%
  select(-c(case_rate_100k,
         avg_humidity_relative_2m_pct,
         avg_particulate_matter_2p5_ugpm3_x10))

```


We can of course use *select()* to *keep* numerous variables we want. Why don't I use a c() operator, below?

```{r}

data4 <- data %>%
  select(fips,
         mon_year,
         avg_temperature_air_2m_f,
         tot_precipitation_in)

head(data4)
```


The reason I used a c() for the first example, when we used a negative, is basically because I didn't want to type a - symbol before each variable name. I could have just as easily done:

```{r}

data3 <- data %>% 
  select(-case_rate_100k,
         -avg_humidity_relative_2m_pct,
         -avg_particulate_matter_2p5_ugpm3_x10)

```

Instead of putting a negative symbol before each variable, I essentially told R "here is a list of variables and I don't want to keep any of them" by putting the negative before the c(). 


Okay, now that we've played with *select()*, let's see what else we did on Day 2 during the live coding session:


Take a look at the original "data" object. See if there are any missing values (NA) or if anything looks suspicious.

```{r}

summary(data)

```


You can also look to see if there are any NAs by asking R to find them, Take a look at the code below. What does is.na() do? 


```{r}

#Let's see how many rows of data had missing values by using the nrow() and is.na() functions
nrow(data[is.na(data),])

```


Zero! Fabulous. No missing data. 

One more feature that we played with was the *rename()* function. We can rename the variables so they are easier to code with:

```{r}

data <- data %>%
  rename("temperature" = avg_temperature_air_2m_f, 
         "precipitation" = tot_precipitation_in, 
         "pm2p5" = avg_particulate_matter_2p5_ugpm3_x10, 
         "wind_speed" = avg_wind_speed_10m_mph,
         "humidity" = avg_humidity_relative_2m_pct)

#look at the data set:
head(data)
```


We finished Day 2 with some plotting. 


That's all for review. Now, let's take a look at our data and make sure we have everything we need to do the next step of exploratory data analysis: bivariate models.

For our analysis, we are not just using our explanatory variables at one point in time. We are also interested in lagged variables: predictors from the past that predict cases now.  Let's create some lagged variables:

```{r}

data <- data %>% group_by(fips) %>%
  arrange(mon_year) %>%
  mutate(
    lag_temperature = lag(temperature, 1),
    lag_precipitation = lag(precipitation, 1),
    lag_pm2p5 = lag(pm2p5, 1),
    lag_humidity = lag(humidity, 1),
    lag_wind_speed = lag(wind_speed, 1))
```


Now that we have all our data, we can run some simple linear models to look at relationships. 



```{r}
# Model: case_rate ~ temperature
model_temp <- lm(case_rate_100k ~ temperature, data = data)
summary(model_temp)

# Model: case_rate ~ precipitation
model_precip <- lm(case_rate_100k ~ precipitation, data = data)
summary(model_precip)

# Model: case_rate ~ pm2p5
model_pm2p5 <- lm(case_rate_100k ~ wind_speed, data = data)
summary(model_pm2p5)
```


Before we move on, let's make sure our final data set is saved in our folder for next steps. You have heard of read.csv(). Now we will write a .csv.

Below, I have commented out the code by putting a # in front of it. This tells R to ignore that row. It's okay to ignore it, because I have already saved final_data.csv in the Day 3 folder. 

```{r}

#summary(data)

#write.csv(data, file = "final_data.csv", row.names = F) #this will save final_data.csv into your working directory.

```

Congratulations! You have completed the day 2 review AND lagged some variables. 